Ripple Protocol Consensus Algorithm Review

### 2015-05-11 paper.tex
checkpoint
close to finished

\documentclass{article}
\usepackage{url}
% \usepackage{showframe}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{multirow}
\usetikzlibrary{arrows,positioning}
\begin{document}
\title{Ripple Consensus Review}
\author{Peter Todd}
\date{FIXME}
\maketitle

\section{Activities performed}

@@

    \item Reviewed whitepapers and development documentation at ripple.com

    \item Reviewed existing third-party critism.

@@

    \item Reviewed C++ implementation, specifically the 0.27.4 tag.\footnote{git commit 92812fe7239ffa3ba91649b2ece1e892b866ec2a}

@@

\section{Overall architecture}

Ripple has diverged significantly from the original concept of a decentralized
network representing money explicitly as debt relationship between
parties.\cite{btcmag-introducing-ripple} While the concept of a debt
relationship still exists in the form of trust lines between participants, the
bulk of the codebase and developer documentation now focuses\footnote{For
    instance the ``Tutorials'' section of the Ripple website only explains how
    to create transactions that modify the ledger; there is almost no information
available on how to actually use the trustlines feature.} on the use and
maintenance of a global ledger of transactions and account balances.
Additionally a native currency, XRP, has been added, which is used for
anti-spam transaction fees\footnote{A small amount of XRP is irrovocably
destroyed for every transaction.}, as well as to serve as a universal currency.

\subsection{Hashing and serialization}

@@

\begin{equation}
    \textit{Serialize}(\text{obj}) = t_0 n_0 d_0 + \hdots + t_n n_n d_n
\end{equation}

Hashing of objects generally uses the following scheme, implemented by
STObject::getHash(), resulting in a $256\text{bit}$ digest:

@@

Unlike Bitcoin, transactions increment and decrement account balances; they do
not directly consume the output of other transactions. To prevent replay
attacks, accounts and transactions have sequence numbers; a transaction is only
valid if the Sequence number is exactly one greater than the last-valided
transaction from the same account. Additionally an optional AccountTxnID field
is available for use within transactions; the transaction is only valid if the
sending account's previously-sent transaction matches the provided hash.

\subsection{Ledger}

Information on the exact structure of the Ripple ledger is somewhat spotty.

how exactly does the skiplist work?

@@

The actual process of determining what version of the ledger is correct starts
with the Unique Node List, (UNL) a list of public keys meant to be associated
with active nodes the node operator believes are
``unique''.\cite{ripple-wiki-unl} Ripple Labs suggests that UNL's ``should have
100+ nodes on them.''\cite{ripple-wiki-unl} and provides a ``starter'' UNL at
\url{https://ripple.com/ripple.txt} with $5$ to $8$ nodes.\footnote{Exactly how this UNL is
generated is unknown; the author downloaded it on multiple occasions getting
between $5$ to $8$ nodes each time, mostly the same.}

Through the Ripple Protocol Consensus Algorithm\cite{ripple-consensus-paper}
nodes on the UNL vote to determine the contents of the Ripple ledger. While the
actual protocol contains a number of rounds of proposals and voting the end
result can be described as basically a supermajority vote: a transaction is
only approved if $80\%$ of the UNL of a server agrees with
it.\cite[3.2]{ripple-consensus-paper} Put another way, if $20\%$ of the UNL
choose to reject a transaction, it will not be included in the ledger.

@@

A closely related issue is what incentive does a node operator have to publicly
perform validation services? Ripple currently has no compensation mechanism for
public validators, yet validation at minimum raises potential legal issues such
as lawsuits for negligence. Again, the least-risk option is to not publicly
validate the Ripple ledger.

@@

The attacker exploits a difference in behavior between different
implementations/versions of the Ripple protocol. The result of the attack can
be a simple denial of service due to the Ripple network being unable to process
transactions, or with more sophistication, the attacker can fork the Ripple
ledger/blockchain and use the fork to doublespend. The attack is stopped by
identifying it, and then distributing software patches. Provided the Ripple
network is well monitored consensus split attacks can be stopped fairly
quickly; the March 2013 Bitcoin chain fork\cite{bip50} was resolved in a matter
of hours by contacting mining pools. A similar response could happen on Ripple
by contacting validating nodes commonly used in UNLs.\footnote{Another reason
to use the UNL provided Ripple Labs.}

@@

    \item The main implementation of the Ripple protocol, \emph{rippled}, does
        not cleanly separate the consensus-critical part of the codebase from
        the non-consensus-critical part.\footnote{The consensus-critical
            portions of Bitcoin Core codebase are currently being separated
            into a stand-alone \emph{libconsensus} library.}

@@

    \item The rippled node implementation does not check full history. This
        means if validators sign an invalid ledger, the transactions that make
        it invalid can be hidden from Ripple users. Equally, even with
        detection, the fact that Ripple users can be asked to simply
        restart/reinstall their nodes to get past the invalid history may
        significantly reduce the social barriers to creating an invalid ledger.

@@

The attacker inserts a backdoor into a widely used Ripple protocol
implementation. (e.g. the rippled node software) The attacker may be an
internal Ripple Labs employee, an external member of the public submitting a
pull-request, or they may compromise the software actually downloaded by
end-users by compromising binary or sourcecode hosting. (e.g. GitHub)

@@

    \item Ripple Labs appears to use GitHub - a third party -
        as their master sourcecode repository, with development happening
        through pull-requests on GitHub. While this does place trust in a
        third-party, PGP signing sourcecode during development can mitigate
        this risk. However there appears to be no PGP signatures on any git
        commits, giving Ripple Labs little ability to audit who wrote what
        code.

@@

    \item The rippled implementation doesn't yet appear to support separating
        the validation/consensus and signing functionality into different
        machines, or even different user accounts. (possibly Ripple Labs has
        internal-use-only improvements)

@@

What's actually happened is Russian cyberwarfare experts created a fake
consensus using stolen Swiss and French private keys. Russia of course
officially signed this fake consensus, as well as the real consensus, taking
the position this was an above-the-board court ordered operation; China also
signed, but doesn't officially say why. The wealthy Chinese investor
conveniently claims to have suddenly found out about the ``fake'' oil importer
and on the real Ripple ledger has the funds frozen by the Chinese bank backing
them - not a big deal with the money still sitting in the original account he
deposited it too, still under Chinese jurisdiction.

@@

>>>>

\documentclass{article}
\usepackage{url}
% \usepackage{showframe}
\usepackage{mathtools}
\usepackage{multirow}
\begin{document}
\title{Ripple Protocol Consensus Algorithm Review}
\author{Peter Todd}
\date{May 11th 2015}
\maketitle

\section{Activities Performed}

@@

    \item Reviewed whitepapers and development documentation at \url{https://ripple.com}.

    \item Reviewed existing third-party criticism.

@@

    \item Reviewed C++ implementation, specifically the 0.27.4 tag of
        \texttt{rippled}.\footnote{git commit
    \texttt{92812fe7239ffa3ba91649b2ece1e892b866ec2a} from \url{https://github.com/ripple/rippled}}

@@

\section{Overall Architecture}

Ripple has diverged significantly from the original
concept\cite{btcmag-introducing-ripple} of a decentralized network recording
the transferral of money explicitly as debt relationships between parties.
(Figure \ref{fig:orig-ripple}, showing entities A, B, and C participating in a
transaction) While the concept of a debt relationship still exists in the form
of trustlines between participants, the bulk of the codebase and developer
documentation now focuses\footnote{For instance the ``Tutorials'' section of
the Ripple website only explains how to create transactions that modify the
ledger; there is almost no information available on how to actually use the
trustlines feature.} on the use and maintenance of a global ledger of
transactions and account balances.  Additionally a native currency, XRP, has
been added, which is used for anti-spam transaction fees\footnote{A small
amount of XRP is irrovocably destroyed for every transaction.} and to
serve as a universal currency. (Figure \ref{fig:ripple-labs-ripple})

\begin{figure}
    \centering
    \includegraphics[scale=0.3]{figures/orig-ripple.eps}
    \caption{Original Ripple Architecture}
    \label{fig:orig-ripple}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale=0.3]{figures/ripple-labs-ripple.eps}
    \caption{Ripple Labs Architecture}
    \label{fig:ripple-labs-ripple}
\end{figure}

\section{Global Ledger}

Figure \ref{fig:ledger-blockchain} shows the basic cryptographic structure of
the Ripple ledger. As with Bitcoin, it consists of a set of chain of blocks,
each consisting of a block header that commits to a previous block and a set of
new transactions for that block. (via a merkle tree) Additionally each block
header also commits to a merkelized trie\footnote{Oddly the trie is not binary,
trading off performance for significantly larger proofs.} of the state of all
account balances. The block chain doesn't itself deal directly with the
consensus protocol. Rather validators sign messages as part of the consensus
protocol, and knowledge of those messages lets Ripple nodes determine the
correct chain.

Developer documentation about exactly how the ledger is structured is somewhat
spotty; the descriptions here are mostly based on direct analysis of the
\texttt{rippled} source code.

\begin{figure}
    \centering
    \includegraphics[scale=0.3]{figures/ledger-blockchain.eps}
    \caption{Ripple Ledger Structure}
    \label{fig:ledger-blockchain}
\end{figure}

Unlike Bitcoin, transactions increment and decrement account balances; they do
not directly consume the output of other transactions. Transactions are also
used to setup and modify trustlines. To prevent replay attacks, accounts and
transactions have sequence numbers; a transaction is only valid if the Sequence
number is exactly one greater than the last-valided transaction from the same
account. Additionally an optional AccountTxnID field is available for use
within transactions; the transaction is only valid if the sending account's
previously-sent transaction matches the provided hash.

\subsection{Low-Level Hashing and Serialization}

@@

\begin{equation}
    \textit{Serialize}(\text{obj}) = t_0+n_0+d_0 + \hdots + t_n+n_n+d_n
\end{equation}

Hashing of objects generally uses the following scheme, implemented by
\texttt{STObject::getHash()}, resulting in a $256\text{bit}$ digest:

@@

\subsection{Open Questions}

    \item How exactly does the block header skiplist work? While the source code
        refers to a skiplist, the actual block header format appears to only
        allow block headers to refer to a single previous block header in a
        linear fashion.

    \item The \texttt{rippled} codebase suggests that in some circumstances account
        balances can be negative. What are those circumstances exactly?

    \item Do opportunities exist to ``shard'' the Ripple blockchain to improve
        scaling by distributing load across parallel servers?

    \item What plans exist (if any) to support SPV-style ``lite clients'' in
        the Ripple protocol?

\section{Consensus Algorithm}

Coming to consensus about the contents of the Ripple ledger is the role of the
Ripple Protocol Consensus Algorithm\cite{ripple-consensus-paper}. In situations
where the participants can be determined in advance, consensus algorithms are a
well studied with a large variety of choices available. With that in mind while
this review did some basic ``sanity checking'' of the specific algorithm, for
the most part we assume it works correctly as described. Instead we'll focus on
the assumptions surrounding the consensus algorithm, and how it can and will be
used in practice.

@@

Like the vast majority of consensus algorithms, the Ripple consensus algorithm
starts with a known set of nodes known to be participating in the consensus.
This Unique Node List, (UNL) is a list of public keys meant to be associated
with active (validating) nodes the node operator believes are
``unique''.\cite{ripple-wiki-unl} Ripple Labs suggests that UNL's ``should have
100+ nodes on them.''\cite{ripple-wiki-unl} and provides a ``starter'' UNL at
\url{https://ripple.com/ripple.txt} with $5$ to $8$ nodes.\footnote{Exactly how
this UNL is generated is unknown; the author downloaded it on multiple
occasions getting between $5$ to $8$ nodes each time, mostly the same.}

Through the consensus algorithm nodes on the UNL vote to determine the contents
of the Ripple ledger. While the actual protocol contains a number of rounds of
proposals and voting the end result can be described as basically a
supermajority vote: a transaction is only approved if $80\%$ of the UNL of a
server agrees with it.\cite[3.2]{ripple-consensus-paper} Put another way, if
$20\%$ of the UNL choose to reject a transaction, it will not be included in
the ledger.

@@

A closely related issue is what incentive does a node operator have to publicly
perform validation services? Ripple currently has no compensation mechanism for
public validators, yet validation at minimum raises potential legal issues such
as lawsuits for negligence or aiding financial crimes. Again, the least-risk
option is to not publicly validate the Ripple ledger.

@@

The attacker exploits a difference in behavior between different
implementations/versions of the Ripple protocol. The result of the attack can
be a simple denial of service due to the Ripple network being unable to process
transactions, or with more sophistication, the attacker can fork the Ripple
ledger/blockchain and use the fork to double-spend. The attack is stopped by
identifying it, and then distributing software patches. Provided the Ripple
network is well monitored consensus split attacks can be stopped fairly
quickly; the March 2013 Bitcoin chain fork\cite{bip50} was resolved in a matter
of hours by contacting mining pools. A similar response could happen on Ripple
by contacting validating nodes commonly used in UNLs.\footnote{Another reason
to use the UNL provided Ripple Labs.}

@@

    \item The main implementation of the Ripple protocol, \texttt{rippled}, does
        not cleanly separate the consensus-critical part of the codebase from
        the non-consensus-critical part.\footnote{The consensus-critical
            portions of Bitcoin Core codebase are currently being separated
            into a stand-alone \emph{libconsensus} library.}

@@

    \item The \texttt{rippled} node implementation does not check full history. This
        means if validators sign an invalid ledger, the transactions that make
        it invalid can be hidden from Ripple users. Equally, even with
        detection, the fact that Ripple users can be asked to simply
        restart/reinstall their nodes to get past the invalid history may
        significantly reduce the social barriers to creating an invalid ledger.

@@

The attacker inserts a backdoor into a widely used Ripple protocol
implementation. (e.g. the \texttt{rippled} node software) The attacker may be an
internal Ripple Labs employee, an external member of the public submitting a
pull-request, or they may compromise the software actually downloaded by
end-users by compromising binary or source code hosting. (e.g. GitHub)

@@

    \item Ripple Labs appears to use GitHub - a third party -
        as their master source code repository, with development happening
        through pull-requests on GitHub. While this does place trust in a
        third-party, PGP signing source code during development can mitigate
        this risk. However there appears to be no PGP signatures on any git
        commits, giving Ripple Labs little ability to audit who wrote what
        code.

@@

    \item The \texttt{rippled} implementation doesn't yet appear to support separating
        the validation/consensus and signing functionality into different
        machines, or even different user accounts. (possibly Ripple Labs has
        internal-use-only improvements)

@@

What has actually happened is Russian cyberwarfare experts created a fake
consensus using stolen Swiss and French private keys. Russia of course
officially signed this fake consensus, as well as the real consensus, taking
the position this was an above-the-board court ordered operation; China also
signed, but doesn't officially say why. The wealthy Chinese investor
conveniently claims to have suddenly found out about the ``fake'' oil importer
and on the real Ripple ledger has the funds frozen by the Chinese bank backing
them - not a big deal with the money still sitting in the original account he
deposited it too, still under Chinese jurisdiction.

@@

\section{Conclusions and future work}

Nearly all the attack scenarios analysed above are direct outcomes of the need
for consensus; the actual blockchain technology that records the ledger is
relatively uninteresting. That the consensus requirement is for global
consensus rather than the original Ripple's local consensus creates additional
problems such as co-ordination of UNL's, incentives for such co-ordination, and
unclear incentives for validating nodes to validate correctly or honestly.
Finally the requirement of global consensus creates scalability concerns,
privacy concerns, and jurisdictional concerns.

A key question that should be answered in future work is if the goals of the
Ripple system need global consensus at all? If global consensus can be avoided,
or at least it's use minimized, many of these issues may go away.
